\chapter{Flow Control}
There are four things you will learn in this chapter. Switching the path of 
execution in program depending upon program variables or states using control 
statements, repeating a set of instructions using loops, bypassing certain set
of instructions in a loop and jump around. Collectively, these elements of C
allow or enable you to take driver's seat over the control over a C program.
You will spend much of your programming time even in future using these basic
elements. It is critical to understand the these topics well as these are basic
pillars over which rest of chapters will build upon.

Let us start with if-else statement \S(6.8.4.1) which is part of selection
statements \S(6.8.4).

\section{if else-if else Statements}
An if statement can be broken into three distinct part. It starts with a
mandatory single \texttt{if} clause which tests an expression and if that expression
evaluates to boolean true then an associated block of code is executed. The
\texttt{if} part may be followed by zero or more `else if` statements which also test
an expression and it can have an associated block of code as well. Finally it
can have an \texttt{else} statement which is optional and does not have any expression
to test against. Rather if all above statements did not match their expressions
then else block's code will be executed. Note that among all blocks of code of
\texttt{if, else if} and \texttt{else} only one block of code will execute and rest will
not.

Let us see a small program to see these in action:

\begin{minted}[frame=single]{c}
#include <stdio.h>

int main()
{
  int i = 0, j= 0;
    
  printf("Please enter two integers i and j:\n");
  scanf("%d%d", &i , &j);
  
  if(i==4)
    printf("you entered 4 for i.\n");

  if(i==7)
  {
    printf("you entered 7 for i.\n");
    printf("I am happy for you.\n");
  }
  else
  {
    printf("You did not enter 7 for i.\n");
  }
  
  if(i==7)
  {
    printf("you entered 7 for i.\n");
    printf("I am happy for you.\n");
  }
  else if(j==8)
    printf("You entered 8 for i.\n");
  
  if(i==7)
    printf("you entered my lucky number.\n");
  else if((i==7) &&(j==8))
    printf("May god bless you!\n");
  else
    printf("You entered bad number.\n");
  
  return 0;
}
\end{minted}

and the output is:
\\\\\texttt{Please enter two integers i and j:\\
4\\
6\\
you entered 4 for i.\\
You did not enter 7 for i.\\
You entered bad number.\\\\}
As you can see from first if sttatement that if you enter the value of \texttt{i} as 4
then the \texttt{printf} will be executed and you will be able to see it. Note that if
there are multiple lines below if which you want to execute then you must put
them in a block using curly braces. If you just want to execute one line then
these curly braces are optional. Note that how you must use curly braces if you
have more than one line and you want to execute that block. Also, see the
syntax for \texttt{else} and \texttt{else if}. One if-else can be nested inside
another for example see the following code:

\begin{minted}[frame=single]{c}
#include <stdio.h>
#include <string.h>
 
int main()
{
  char fName[128]={0}, lName[128]={0};
 
  printf("Enter your first name and last name in that order:\n");
  gets(fName);
  gets(lName);
 
  if(strcmp(fName, "Shiv") == 0)
  {
    if(strcmp(lName, "Dayal") == 0)
      printf("Your name is Shiv Dayal.\n");
  }
  else
  {
    printf("Your name is %s %s.\n", fName, lName);
  }
 
  return 0;
}
\end{minted}

and the output is:
\\\\\texttt{Enter your first name and last name in that order:\\
Shiv\\
Dayal\\
Your name is Shiv Dayal.\\\\}
another run when first \texttt{if} fails:
\\\\\texttt{Enter your first name and last name in that order:\\
Richard\\
Stallman\\
Your name is Richard Stallman.\\\\}
when first \texttt{if} matches but second \texttt{if} does not and this we have no output:
\\\\\texttt{Enter your first name and last name in that order:\\
Shiv\\
Stallman\\\\}
Note the usage of nested if-else. Also, note how \texttt{strcmp} has been used to
compare two strings and \texttt{gets} to read the input. \texttt{strcmp} takes two character
strings as argument and returns 0 if they are equal. It returns non-zero values
depending on whether one string is lexically greater than the other or not. But
for now equality is enough for us. \texttt{gets} is dangerous but it is simple that is
why has been used here. It is easy to overflow the buffer of \texttt{gets} argument.

\color{nicered}
\dangersign[3ex] \textbf{Assignment in if/else-if}
Always remember the expression inside if evaluates to a boolean so you should
never do an ASSIGNMENT inside if and else if as it will always evaluate to what
is assigned. It can render all your logic meaningless. C is not Python, where
assignment inside if is not allowed. However, if you assign 0 to some variable
it will evaluate to \texttt{false}.

\color{black}
\subsection{Dangling else Problem}
The \texttt{else} part has a property that it will cling to closest if. So the
following piece of code may give you surprise:

\begin{minted}[frame=single]{c}
if(x==1)
  if(y>2)
    printf("foo\n");
else
  printf("bar\n");
\end{minted}

Now consider \texttt{x!=1} then you may think that bar will be
printed. However, that will not be the case. The else part clings to inner
if. This can be fixed by using curly braces.

\section{\texttt{switch} Statement}
\texttt{switch} statement is kind of if-else replacement to simplify it. Usage
of switch statement is to compare one expression with others, and then execute
a series of sub-statements inside case and default based on the result of the
comparisons. Note that switch statement takes only integers or integreal type
as its argument and same is valid for its cases. Consider the following
example:

\begin{minted}[frame=single]{c}
//Description : Demo of if-else statements.

#include <stdio.h>

int main()
{
  int i  = 65;

  switch(i)
  {
    case 'A':
      printf("Value of i is 'A'.\n");
      break;
    case 'B':
      printf("Value of i is 'B'.\n");
      break;
    default:
      break;
  }

  return 0;
}
\end{minted}
and the output is:
\\\\\texttt{Value of i is 'A'.\\\\}
Notice the usage of \texttt{break}. It is used to terminate execution once a
match has been found for a particular case else what will happen is shown
below:

\begin{minted}[frame=single]{c}
//Description : Demo of switch statement.

#include <stdio.h>

int main()
{
  int i  = 65;

  switch(i)
  {
    case 'A':
      printf("Value of i is 'A'.\n");
    case 'B':
      printf("Value of i is 'B'.\n");
    default:
      printf("Value of i is %c.\n", i);
      break;
    }

  return 0;
}
\end{minted}
and the output is:
\\\\\texttt{Value of i is 'A'.\\
Value of i is 'B'.\\
Value of i is A.\\\\}
This is also known as fall through of a \texttt{switch} statement. Notice, the
use of \texttt{default} that how it is analogous to else
statement. \texttt{switch} statements can also be nested inside each
other. However, node that lots of nesting is not good. At 
most 2-3 levels are more than enough else you should look at alternative ways
of writing code.

\section{\texttt{while} Loop}
Of three loops I am first going to cover \texttt{while} loop. It is simplest of
three. I will just give an example for you to understand.

\begin{minted}[frame=single]{c}
//Description : Demo of while statement.

#include <stdio.h>

int main()
{
  int i = 0;

  while(i <= 10)
  {
    printf("%d * %2d = %4d\n", 2, i, 2*i);
    i++;
  }

  return 0;
}
\end{minted}
and the output is:
\\\\\texttt{2 *  0 =    0\\
2 *  1 =    2\\
2 *  2 =    4\\
2 *  3 =    6\\
2 *  4 =    8\\
2 *  5 =   10\\
2 *  6 =   12\\
2 *  7 =   14\\
2 *  8 =   16\\
2 *  9 =   18\\
2 * 10 =   20\\\\}
\texttt{while} loop just has one expression which is its terminating
condition. We have written \texttt{i<=10} which is terminating condition for
our loop. The moment i will become greater than that the loop will
terminated. We are initializing our loop index to 0 and incrementing within
while loop. Note that you must use curly braces for body of block of loop. If
you have only one statement as body of loop then braces are optional.

\section{\texttt{do-while} Loop}
It is very much similar to while loop but with a very subtle
difference. Consider the following code:

\begin{minted}[frame=single]{c}
//Description : Demo of do while statement.

#include <stdio.h>

int main()
{
  int i = 0;

  do {
    printf("%d\n", i);
    i++;
  }while(i<5);

  return 0;
}
\end{minted}
and the output is:
\\\\\texttt{0\\
1\\
2\\
3\\
4\\\\}
Notice the semicolon at the end of \texttt{while}. Now time for that subtle
difference:

\begin{minted}[frame=single]{c}
//Description : Demo of do while statement.

#include <stdio.h>

int main()
{
  int i = 10;

  do {
    printf("2 * %d = 20\n", i);
    i++;
  }while(i<5);

  return 0;
}
\end{minted}
and the output is:
\\\\\texttt{2 * 10 = 20\\\\}
Notice how \texttt{do while} loop executes once even if the loop index is more
than the terminating condition in the \texttt{while} part.
\section{\texttt{for} Loop}
\texttt{for} loop is the last of loops and most versatile. It has three parts:
initialization of loop counters, terminating condition, and loop index
modification. If you declare a variable in the initialization part then that
variable has just loop scope while for \texttt{while} and \texttt{do while}
loop indices have at least outer block scope. This makes for loop
better. Consider the following example for computing squares of numbers:

\begin{minted}[frame=single]{c}
//Description : Demo of for statement.

#include <stdio.h>

int main()
{
  for(int i=1, j=1; (i<=10)||(j<=10); i++, j++)
    printf("%2d * %2d = %4d\n", i, j, i*j);

  return 0;
}
\end{minted}
and the output is:
\\\\\texttt{1 *  1 =    1\\
2 *  2 =    4\\
3 *  3 =    9\\
4 *  4 =   16\\
5 *  5 =   25\\
6 *  6 =   36\\
7 *  7 =   49\\
8 *  8 =   64\\
9 *  9 =   81\\
10 * 10 =  100\\\\}
Notice how various things are coming in picture here: initialization,
terminating conditions loop counter incrementation and output formatting. Here
is how you can write an infinite for loop \texttt{for(;;)}. You can write an
infinite loop anywhere if your loop index counters are not getting
incremented/decremented properly or your termination condition is
incorrect. Also, always make sure that loop indices are initialized. As an
exercise you can try to implement this program using \texttt{while} and
\texttt{do while} loop.

\section{\texttt{break} and \texttt{continue} Statements}
\texttt{break} statement breaks out of innermost \texttt{for, do, while} and
\texttt{switch} statements. It terminates that loop. Consider for example:

\begin{minted}[frame=single]{c}
//Description : Demo of break statement.

#include <stdio.h>

int main()
{

  for(int i = 0;;i +=10)
  {
    if(i>100)
      break;
    printf("%d\n", i);
  }

  return 0;
}
\end{minted}
and the output is:
\\\\\texttt{0\\
10\\
20\\
30\\
40\\
50\\
60\\
70\\
80\\
90\\
100\\\\}
Notice how the \texttt{for} loop is terminated once \texttt{i} goes beyond 100
even though there is no terminating condition. Try the same in \texttt{while}
and \texttt{do-while} loop and produce the same result.

\texttt{continue} statement is slightly different than \texttt{break} in the
sense that it does not stop the execution of that loop but simply does not
execute remaining instructions of that block. Consider for example:

\begin{minted}[frame=single]{c}
//Description : Demo of continue statement.

#include <stdio.h>

int main()
{

  for(int i = 0;i<=100;i +=10)
  {
    if(i==50)
      continue;
    printf("%d\n", i);
  }

  return 0;
}
\end{minted}
and the output is:
\\\\\texttt{0\\
10\\
20\\
30\\
40\\
60\\
70\\
80\\
90\\
100\\\\}
Notice how 50 is missing from output.

\section{\texttt{goto} Statement}
\texttt{goto} statement allows you to jump to a label within a function
unconditionally. This leads to arbitrary control flow and in a big function you
can loose track where the code is leading you to. In fact many coding standards
forbid the usage of \texttt{goto} statement. Sometimes you can use to break out
of several level of nested loops but you can use certain techniques to come out
of nested loops to break out of them. Consider the following program:

\begin{minted}[frame=single]{c}
#include <stdio.h>

int main()
{
  int i = 0;
  
  goto EXIT;

  printf("Will not be printed\n");
  
EXIT:
  printf("This is where goto will lead you to.\n");

  return 0;
}
\end{minted}
and the output is:
\\\\\texttt{This is where goto will lead you to.\\\\}
One of the usage of \texttt{goto} is to simulate the functionality of
loops. That is easy to understand because a CPU does not have any instructions
for looping. Rather, loop statements are translated to comparison and jump
instructions.

Consider we have two nested \texttt{for} loops which run from 0 to 9 and we
want to break out when outer loop's counter is 5 and inner loop's counter is 7
the we can use \texttt{goto} as given below:

\begin{minted}[frame=single]{c}
#include <stdio.h>

int main()
{
  for(int i=0; i<10; ++i)
  {
    for(int j=0; j<10; ++j)
    {
      if((i == 5) && (j == 7))
      {
        printf("i = %d and j = %d\n", i, j);
        goto EXIT;
      }
    }
  }

EXIT:

  return 0;
}
\end{minted}
and the output is:
\\\\\texttt{i = 5 and j = 7\\\\}
However, it is possible to break out of such nested loops using a flag variable
as shown below:

\begin{minted}[frame=single]{c}
#include <stdio.h>
#include <stdbool.h>

int main()
{
  bool flag = false;

  for(int i=0; i<10; ++i)
  {
    for(int j=0; j<10; ++j)
    {
      printf("i = %d and j = %d\n", i, j);
      if((i == 5) && (j == 7))
      {
        printf("i = %d and j = %d\n", i, j);
        flag = true;
        break;
      }
    }
    if(flag)
      break;
  }

  return 0;
}
\end{minted}
and the output is:
\\\\\texttt{i = 0 and j = 0\\
i = 0 and j = 1\\
i = 0 and j = 2\\
i = 0 and j = 3\\
i = 0 and j = 4\\
i = 0 and j = 5\\
i = 0 and j = 6\\
i = 0 and j = 7\\
i = 0 and j = 8\\
i = 0 and j = 9\\
i = 1 and j = 0\\
i = 1 and j = 1\\
i = 1 and j = 2\\
...\\
i = 5 and j = 1\\
i = 5 and j = 2\\
i = 5 and j = 3\\
i = 5 and j = 4\\
i = 5 and j = 5\\
i = 5 and j = 6\\
i = 5 and j = 7\\
i = 5 and j = 7\\\\}
This simple technique can be used to break out of several levels of nested
loops.

\section{Examples}
Now that we have studied control flow and operators and expressions we can
write simple but very interesting programs. Given below are few examples.

\subsection{Implementing a Loop Using \texttt{goto} Statement}
As said above a CPU typically does not have loop instructions but a loop is
translated into comparison, increment/decrement and jump instructions. Thus, a
loop can be implemented using \texttt{goto, if} and increment/decrement
statements and operators. Consider the following program which prints 1 to 10.

\begin{minted}[frame=single]{c}
#include <stdio.h>

int main()
{
  for(int i=1; i<11; ++i)
    printf("%d\n", i);

  return 0;
}
\end{minted}
and the output is:
\\\\\texttt{1\\
2\\
3\\
4\\
5\\
6\\
7\\
8\\
9\\
10\\\\}
and the equivalent simulation using \texttt{goto} is:

\begin{minted}[frame=single]{c}
#include <stdio.h>

int main()
{
  int i = 1;

LOOP:
  printf("%d\n", i);
  ++i;
  if(i!= 11)
    goto LOOP;

  return 0;
}
\end{minted}
and the output is same as above which you can easily verify.

\subsection{Printing Various Patterns}
Consider we want to print following pattern:
and the output is:
\begin{verbatim}
    *
   ***
  *****
 *******
*********
\end{verbatim}
 
then how would you print it. It is a very easy program and I will just give you
the solution. First try to do it yourself and use the solution if you cannot do
it to learn from it.

\begin{minted}[frame=single]{c}
#include <stdio.h>
 
int main()
{
  unsigned int c = 0, n = 0, temp = 0;
 
  printf("Enter the number of rows in the pattern: ");
  scanf("%u",&n);
 
  temp = n;
 
  for (int row = 1; row <= n; row++)
  {
    for (c = 1; c < temp; c++)
      printf(" ");
 
    temp--;
 
    for ( c = 1 ; c <= 2*row - 1 ; c++ )
      printf("*");
 
    printf("\n");
  }
 
   return 0;
}
\end{minted}
and the output is:

\begin{verbatim}
Enter the number of rows in the pattern: 10
         *
        ***
       *****
      *******
     *********
    ***********
   *************
  ***************
 *****************
*******************
\end{verbatim}
Another simple pattern may look like:

\begin{verbatim}
*
**
***
****
*****
******
\end{verbatim}
This is even easier than previous one. Try to do it on your own before seeing
the solution given below:

\begin{minted}[frame=single]{c}
#include <stdio.h>
 
int main()
{
  int n=0;
 
  printf("Enter number of rows\n");
  scanf("%d",&n);
 
  for(int c=1 ; c<=n; c++)
  {
    for(int k=1 ; k<=c ; k++ )
      printf("*");
 
    printf("\n");
  }
  
  return 0;
}
\end{minted}
and the output is:
\\\\\texttt{Enter number of rows\\
10\\
*\\
**\\
***\\
****\\
*****\\
******\\
*******\\
********\\
*********\\
**********\\\\}

\subsection{Printing Pascal's Triangle}
Pascal's triangle is discussed in great detail at
\url{https://en.wikipedia.org/?title=Pascal\%27s\_triangle}. To start with let
us look at a small Pascal triangle.

\begin{verbatim}
   1
  1 1
 1 2 1
1 3 3 1
\end{verbatim}

The rows in Pascal's triangle start at 0(say we use $n$ to denote it) and
columns start at 0 as well. We can pick either left end or right end that does
not matter because it is symmetric at the center. Since we are used to writing
left-to-right in English let us pick the first column is at left most end is
say $k$ starting from 0. Now the number at any position is ${}_{}^nC_k^{}$ in
Pascal's triangle. However, it is not necessary to compute each and every
element. As we know, ${}_{}^nC_0^{}$ and ${}_{}^nC_n^{}$ are 1 each thus both
the edges will always be 1. So for 0th row ${}_{}^0C_0^{}$ is the only element
while for the first row it is ${}_{}^1C_0^{}$ and ${}_{}^1C_1^{}$ which both
evaluates to 1 thus 1st row is 1 1. Now from basics of permutation and
combination it follows that ${}_{}^nC_k^{} = {}_{}^{n-1}C_{k-1}^{} +
{}_{}^{n-1}C_k^{}$. Thus, new elements of a new line except for the edges can
be figured from the previous line. For example, 2nd row will contain 3 elements
with two edge-elements being 1. Now the middle element is ${}_{}^2C_1^{}$ which
can be simply presented as sum of ${}_{}^1C_0^{} + {}_{}^1C_1^{}$ which are two
elements of 1st row. Now to computer 3rd row elements we can follow the same
process. The edge elements will be 1 each while the other two elements are
${}_{}^3C_1^{}$ and ${}_{}^3C_2^{}$. Now, ${}_{}^3C_1^{}$ can be obtained by
adding ${}_{}^2C_0^{}$ and ${}_{}^2C_1^{}$ i.e. 1 and 2 respectively, which
result in 3. Similarly, the next element will also result in 3. Thus 3rd row is
1 3 3 1. Similarly, we can computer the next row with ease which will be 1 4 6
4 1 and next row to that will be 1 5 10 10 5 1. Let us write a simple program
to print Pascal's triangle based on these observations:

\begin{minted}[frame=single]{c}
#include <stdio.h>
 
int main()
{
  unsigned int n = 0;
 
  printf("Enter the number of rows in the Pascal's triangle: ");
  scanf("%u",&n);
 
  unsigned long a[n], b[n];
  
  a[0] = b[0] = b[1] = 1;

  printf("%lu\n", a[0]);
  printf("%lu %lu\n", b[0], b[1]);

  unsigned long *x = a, *y = b;
  unsigned long *temp;
  
  for(unsigned long i = 2; i<n; ++i) {
    // set edge values
    x[0] = 1;
    x[i] = 1;
    printf("%lu ", x[0]);
    for(unsigned long j = 1; j<i; ++j) {
      x[j] = y[j - 1] + y[j];
      printf("%lu ", x[j]);
    }
    printf("%lu\n", x[i]);

    // swap pointers for arrays
    temp = x;
    x = y;
    y = temp;
  }
  return 0;
}
\end{minted}
I have not tried to make the output beautiful because with changing input the
output has to adjusted. Thus the formatting changes will destroy the beauty of
this simple program. The output may look like below:
\\\\\texttt{1\\
1 1\\
1 2 1\\
1 3 3 1\\
1 4 6 4 1\\
1 5 10 10 5 1\\
1 6 15 20 15 6 1\\
1 7 21 35 35 21 7 1\\
1 8 28 56 70 56 28 8 1\\
1 9 36 84 126 126 84 36 9 1\\\\}

\section{Converting Decimal Numbers to Binary Numbers}
Since we deal mostly with binary at least at a lower level if not in
applications thus it is essential that we understand how to convert decimal to
binary. We have already seen how to convert decimal to binary in
\autoref{bns} \nameref{bns} so
let us see how to convert a number to binary. Since C does not have a data type
for binaries neither you can use a conversion-specifier to print numbers in
binary I have used C style strings discussed later in the book to convert
source number which is accepted as string in program. Since the input number is
a string it is slightly complicated to convert but then it allows you to deal
with much larger number than supported by \texttt{unsigned long long} data
type. This program uses arrays and functions discussed elsewhere in book so if
you do not understand then skip it and come back later. Here is the program:

\begin{minted}[frame=single]{c}
#include <stdio.h>
#include <string.h>

const unsigned short int MAX = 128;

int main()
{
  char input[MAX];
  char temp[MAX];
  char output[4*MAX]; // A single digit 9 is 4 bits while 99 is 7 bits so 
                      // 4*MAX should fit the converted decimal in bits

  memset(input, 0, MAX);
  memset(temp, 0, MAX);
  memset(output, 0, 4*MAX);

  // read input as string
  printf("Enter a decimal number to be converted to integer:\n");
  fgets(input, 128, stdin);

  // substitute last '\n' from keyboard input with '\0'
  unsigned short int length = strlen(input) - 1;
  input[length] = 0;
  
  // input validation. in ASCII '0' to '9' occur in sequence.
  for(unsigned short int i=0, j=0; i<length; ++i, ++j)
  {
    temp[j] = 0;

    if((input[i]<'0') || (input[i]>'9'))
    {
      printf("Invalid input.\n");
      return -1;
    }
  }

  unsigned short int rem = 0;
  unsigned short int quo = 0;  
  unsigned short int j = 0;

  if((strcmp(input, "0") == 0) || strcmp(input, "1") == 0)
  {
    puts(input);
    return 0;
  }

  while(strcmp(input, "1") != 0)
  {
    // division by 2.
    for(unsigned short int i=0, k=0; i<length; ++i, ++k)
    {
      // 48 is decimal value of '0' so we subtract 48
      // from characters to get numbers
      unsigned short int si = input[i] - 48;
    
      if(rem != 0)
        si += 10*rem;
    
      if(si == 1 && (i+1)<length) {
        si = 10;
        si += input[i + 1] - 48;
        i = i + 1;
        if(k != 0) {
          temp[k] = 48;
          k += 1;
        }
      }

      rem = si%2;
      quo = si/2;

      temp[k] = quo + 48;
    }
    strcpy(input, temp);
    memset(temp, 0, MAX);

    length = strlen(input);
    output[j++] = rem + 48;
    rem = 0;
  }

  output[j] = quo + 48;
  length = strlen(output);

  // reverse output
  for(unsigned short int i=0; i<(length/2); ++i)
  {
    output[i] ^= output[length - i - 1];
    output[length - i - 1] ^= output[i];
    output[i] ^= output[length - i -1];
  }
  
  puts(output);

  return 0;
}
\end{minted}

 The program is commented so as to simplify the understanding. Try to run it and see the output. 
